# Joint and Conditional Distributions {#chap-joint-conditional}

Up to this point, our exploration of probability has focused on analyzing a single random variable at a time. This approach is useful for building a foundation.  In practice, if we are going to the trouble to make any measurements we will make multiple measurements.  Thinking of variable as another word for measurement, multiple measurements means multivariate. Whether we’re studying blood pressure and cholesterol levels in a medical study, rainfall and crop yield in agricultural research, or revenue and customer ratings in a business context, it is rarely sufficient to consider each variable in isolation.

Relationships among variables can be complex, and the picture painted by their joint behavior may differ substantially from the pictures painted by considering them individually (which will be called **marginally**). In this chapter, we extend our framework to accommodate **multivariate random variables**. We will develop tools for analyzing how variables interact with one another, including **expected values**, **[covariance]{#def-joint-and-conditional-distributions-covariance}**, and **[correlation]{#def-joint-and-conditional-distributions-correlation}**, as well as methods for isolating the behavior of one variable given the value of another — a concept known as **conditional probability**.

## Jointly Distributed Random Variables

### Discrete Random Variables

To begin our study of multivariate probability, we consider the case of two discrete random variables. We can describe their joint behavior using a joint probability mass function (or [joint pmf]{#def-joint-and-conditional-distributions-joint-pmf}). The joint pmf provides the probability that both variables simultaneously take on specific values (hence, an **intersection**) — and serves as the foundation for understanding how the two variables relate to one another.

:::: {.colbox data-latex=""}
::: {.definition #chap-joint-joint-pmf}
<!-- Let $X$ and $Y$ be discrete rvs defined on the same sample space. -->
The **joint probability mass function** (or **joint pmf**) of $X$ and $Y$, denoted $f(x, y)$, is defined as

$$
f_{X,Y}(x, y) = P(X = x \text{ and } Y = y)
$$

for each pair of possible values $x$ and $y$. The joint pmf satisfies the following properties:

- $f_{X,Y}(x, y) \geq 0$ for all $x, y$
- $\sum_x \sum_y f_{X,Y}(x, y) = 1$

<!--We can compute the probability that $(X, Y)$ falls in any subset $A$ of the possible $(x, y)$ pairs by summing over that subset:

$$
P\left( (X, Y) \in A \right) = \sum_{(x, y) \in A} f_{X,Y}(x, y)
$$
-->
:::
:::: 

:::: {.whitebox data-latex=""}
::: {.example #chap-joint-coffee-shop}
A small coffee shop tracks customer behavior during weekday mornings and records two variables for each customer:

- $X$: the number of espresso shots ordered (0, 1, or 2), and  
- $Y$: whether the customer also buys a pastry (0 = no, 1 = yes).

Based on historical data, the joint distribution of $(X, Y)$ is summarized in the following table:

|               | $Y = 0$ (No Pastry) | $Y = 1$ (Pastry) |
|---------------|--------------------|------------------|
| $X = 0$ shots | 0.05               | 0.10             |
| $X = 1$ shot  | 0.20               | 0.25             |
| $X = 2$ shots | 0.15               | 0.25             |

Each entry in the table represents the joint probability $f_{X,Y}(x, y) = P(X = x \text{ and } Y = y)$. For example, $f_{X,Y}(1, 0) = 0.20$ means that 20% of customers order 1 shot of espresso and do *not* buy a pastry.

We can check that this is a valid joint pmf:

\[
\sum_x \sum_y f_{X,Y}(x, y) = 0.05 + 0.10 + 0.20 + 0.25 + 0.15 + 0.25 = 1
\]

Use the joint pmf to compute the probability that a customer orders more than 0 espresso shots *and* also buys a pastry. That is, compute $P(X > 0 \text{ and } Y = 1)$.

_Solution_: We add the relevant joint probabilities:

\[
P(X > 0 \text{ and } Y = 1) = f_{X,Y}(1, 1) + f_{X,Y}(2, 1) = 0.25 + 0.25 = 0.50
\]

So, there is a 50% chance that a customer orders at least one shot of espresso and also buys a pastry.
:::
:::: 

:::: {.colbox data-latex=""}
::: {.definition #chap-joint-marginal-pmf}
Let $X$ and $Y$ be two discrete rvs with joint pmf $f_{X,Y}(x, y) = P(X = x \text{ and } Y = y)$. The **[marginal pmf]{#def-joint-and-conditional-distributions-marginal-pmf}** of $X$ is obtained by summing over all possible values of $Y$:

\[
f_X(x) = \sum_y f_{X,Y}(x, y)
\]

Similarly, the marginal pmf of $Y$ is obtained by summing over all possible values of $X$:

\[
f_Y(y) = \sum_x f_{X,Y}(x, y)
\]

That is, the marginal distribution of one variable describes its probability behavior _regardless of_ the value of the other variable.
:::
:::: 

:::: {.whitebox data-latex=""}
::: {.example #chap-joint-coffee-shop-marginal}
In the context of Example \@ref(exm:chap-joint-coffee-shop), compute the marginal pmfs of $X$ and $Y$ by summing over the rows and columns, respectively.

_Solution_:

(1) Marginal pmf of $X$:

\[
\begin{aligned}
f_X(0) &= f_{X,Y}(0, 0) + f_{X,Y}(0, 1) = 0.05 + 0.10 = 0.15 \\
f_X(1) &= f_{X,Y}(1, 0) + f_{X,Y}(1, 1) = 0.20 + 0.25 = 0.45 \\
f_X(2) &= f_{X,Y}(2, 0) + f_{X,Y}(2, 1) = 0.15 + 0.25 = 0.40
\end{aligned}
\]

(2) Marginal pmf of $Y$:

\[
\begin{aligned}
f_Y(0) &= f_{X,Y}(0, 0) + f_{X,Y}(1, 0) + f_{X,Y}(2, 0) = 0.05 + 0.20 + 0.15 = 0.40 \\
f_Y(1) &= f_{X,Y}(0, 1) + f_{X,Y}(1, 1) + f_{X,Y}(2, 1) = 0.10 + 0.25 + 0.25 = 0.60
\end{aligned}
\]

These marginal distributions describe the individual behavior of $X$ and $Y$ on their own, without reference to the other variable.
:::
:::: 

Notice that a marginal pmf is itself a proper pmf. For example, in the coffee shop context, we have that 
$$
\sum_x f_X(x) = 0.15 + 0.45 + 0.40 = 1
$$
and
$$
\sum_y f_Y(y) = 0.40 + 0.60 = 1
$$

### Continuous Random Variables

Just as we defined a joint probability mass function (pmf) for two discrete rvs, we can define a joint probability density function (pdf) for two continuous rvs. The [joint pdf]{#def-joint-and-conditional-distributions-joint-pdf} characterizes the behavior of the pair $(X, Y)$ and enables us to compute the probability that the two variables fall within a particular region of the plane.

:::: {.colbox data-latex=""}
::: {.definition #chap-joint-joint-pdf}
Let $X$ and $Y$ be continuous rvs defined on the same sample space. The **joint probability density function** (or **joint pdf**) of $X$ and $Y$, denoted $f_{X,Y}(x, y)$, is a function satisfying:

- $f_{X,Y}(x, y) \geq 0$ for all $x$ and $y$
- $\displaystyle \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} f_{X,Y}(x, y) \, dy \, dx = 1$

For any region $A$ in the $xy$-plane, the probability that $(X, Y)$ falls within $A$ is given by:

\[
P\left( (X, Y) \in A \right) = \iint_{A} f(x, y) \, dx \, dy
\]

In particular, for a rectangular region $a \leq X \leq b$, $c \leq Y \leq d$, we have:

\[
P(a \leq X \leq b, \, c \leq Y \leq d) = \int_a^b \int_c^d f(x, y) \, dy \, dx
\]
:::
:::: 

:::: {.whitebox data-latex=””}
::: {.example #chap-joint-continuous-toys}
A toy manufacturing company produces customized building blocks. Let $X$ represent the proportion of blue pigment used in a block’s color formula (on a scale from 0 to 1), and let $Y$ represent the proportion of red pigment; total pigment content cannot exceed 100% (i.e., $x + y \leq 1$). 

The joint behavior of $X$ and $Y$ follows the pdf
$$
f_{X,Y}(x, y) = 
\begin{cases}
24xy, & \text{if } 0 \leq x \leq 1,\ 0 \leq y \leq 1,\ \text{and } 0 \leq x + y \leq 1 \\\\
0, & \text{otherwise}
\end{cases}
$$

(a) Verify that this is a valid pdf.

(b) Compute $P(X > 0.5,\ Y < 0.5)$.

_Solution_:

(a) We verify that the total probability is 1 by integrating over the triangular region defined by $0 \leq x \leq 1$ and $0 \leq y \leq 1 - x$:

\begin{align*}
\int_0^1 \int_0^{1 - x} 24xy\, dy\, dx
&= \int_0^1 24x \left[ \frac{1}{2}y^2 \right]_0^{1 - x} dx 
= \int_0^1 12x(1 - x)^2 dx \\
&= \int_0^1 12x(1 - 2x + x^2)\, dx 
= 12 \left[ \frac{1}{2}x^2 - \frac{2}{3}x^3 + \frac{1}{4}x^4 \right]_0^1 \\
&= 12 \left( \frac{1}{2} - \frac{2}{3} + \frac{1}{4} \right) 
= 12 \cdot \frac{1}{12} = 1
\end{align*}

Thus, $f_{X,Y}$ is a valid pdf.

(b) This probability corresponds to the region $A = \{0.5 \leq x \leq 1 \text{ and } 0 \leq y \leq 0.5\}$.
Additionally, we have the three constraints on the pdf itself: $0 \leq x, y \leq 1$ and  $0 \leq x + y \leq 1$. For $0.5 \leq x \leq 1$, the upper limit of $y$ is $\min(0.5, 1 - x)$, which simplifies to $0 \leq y \leq 1 - x$ (since $1 - x \leq 0.5$ in this range). Therefore:

\begin{align*}
P(X > 0.5,\ Y < 0.5) &= \int_{0.5}^1 \int_0^{1 - x} 24xy\, dy\, dx
= \int_{0.5}^1 24x \left[ \frac{1}{2}y^2 \right]_0^{1 - x} dx \\
&= \int_{0.5}^1 12x(1 - x)^2 dx
= \int_{0.5}^1 (12x - 24x^2 + 12x^3)\, dx \\
&= \left[ 6x^2 - 8x^3 + 3x^4 \right]_{0.5}^1
= \left( 6 - 8 + 3 \right) - \left( 6 \left( \frac{1}{2} \right)^2 - 8 \left( \frac{1}{2} \right)^3 + 3 \left( \frac{1}{2} \right)^4 \right) \\
&= 1 - \left( \frac{3}{2} - 1 + \frac{3}{16} \right) 
= \frac{5}{16} = 0.3125
\end{align*}

Thus, there is a .3125 probability that a randomly selected block uses more than 50% blue pigment and less than 50% red pigment.
:::
::::

Figure \@ref(fig:toy-joint-dist) shows the joint pdf corresponding to Example \@ref(exm:chap-joint-continuous-toys), with the triangular region in the $(x, y)$ plane corresponding the $P(X > 0.5,\ Y < 0.5)$ calculation shaded red.

```{r toy-joint-dist, echo=FALSE, message=FALSE, fig.cap= "Joint pdf f(x,y)=24xy with shaded region $(x>0.5,\\;y<0.5,\\;x+y\\le1)$"}
require(plotly)

# Define the joint PDF
f_xy <- function(x, y) {
  24 * x * y
}

resolution = 200
# Create a grid over the unit triangle
x_vals <- seq(0, 1, length.out = resolution)
y_vals <- seq(0, 1, length.out = resolution)
grid <- expand.grid(x = x_vals, y = y_vals)
grid <- subset(grid, x + y <= 1)

# Compute z values (joint PDF)
grid$z <- f_xy(grid$x, grid$y)

# Create matrix for surface plot
z_matrix <- matrix(NA, nrow = resolution, ncol = resolution)
for (i in 1:nrow(grid)) {
  x_idx <- which.min(abs(x_vals - grid$x[i]))
  y_idx <- which.min(abs(y_vals - grid$y[i]))
  z_matrix[y_idx, x_idx] <- grid$z[i]
}

# Define mask for shaded region: x > 0.5 and y < 0.5
mask <- outer(y_vals, x_vals, function(y, x) (x > 0.5 & y < 0.5 & (x + y <= 1)))

# Plot
plot_ly(showscale = FALSE) %>%
  add_surface(
    x = ~x_vals, y = ~y_vals, z = ~z_matrix,
    colorscale = list(c(0, 1), c("lightblue", "blue")),
    opacity = 0.9,
    name = "Joint PDF"
  ) %>%
  add_trace(
    type = "surface",
    x = ~x_vals,
    y = ~y_vals,
    z = ~ifelse(mask, 0, NA),
    surfacecolor = ~ifelse(mask, 1, NA),
    colorscale = list(c(0, 1), c("red", "red")),
    showscale = FALSE,
    opacity = 0.7,
    name = "Shaded region"
  ) %>%
  layout(
    #title = "Joint PDF f_{X,Y}(x, y) = 24xy with Shaded Region (X > 0.5 & Y < 0.5)",
    scene = list(
      xaxis = list(title = "X"),
      yaxis = list(title = "Y"),
      zaxis = list(title = "f(x, y)")
    )
  )
```

:::: {.colbox data-latex=""}
::: {.definition #chap-joint-marginal-pdf}
Let $X$ and $Y$ be two continuous random variables with joint pdf $f_{X,Y}(x, y)$. The **[marginal pdf]{#def-joint-and-conditional-distributions-marginal-pdf}** of $X$ is defined as

$$
f_X(x) = \int_{-\infty}^{\infty} f(x, y)\, dy
$$

Similarly, the marginal pdf of $Y$ is defined as

$$
f_Y(y) = \int_{-\infty}^{\infty} f(x, y)\, dx
$$

These marginal distributions describe the individual probability behavior of each variable, _regardless of_ the value of the other.
:::
::::

:::: {.whitebox data-latex=””}
::: {.example #chap-joint-toys-marginal}
In the context of Example \@ref(exm:chap-joint-continuous-toys), we defined the joint pdf of $X$ and $Y$ as

$$
f_{X,Y}(x, y) =
\begin{cases}
24xy, & \text{if } 0 \leq x \leq 1,\ 0 \leq y \leq 1,\ \text{and } x + y \leq 1 \\
0, & \text{otherwise}
\end{cases}
$$

Derive the marginal pdfs of $X$ and $Y$.

_Solution_:

(a) Marginal pdf of $X$:

We integrate over the values of $y$ that satisfy the constraint $x + y \leq 1$ (i.e., $0 \leq y \leq 1 - x$):

$$
f_X(x) = \int_0^{1 - x} 24xy dy = 24x \int_0^{1 - x} y dy
= 24x \cdot \left[ \frac{1}{2} y^2 \right]_0^{1 - x}
= 12x(1 - x)^2
$$

for $0 \leq x \leq 1$. Thus,

$$
f_X(x) =
\begin{cases}
12x(1 - x)^2, & 0 \leq x \leq 1 \\
0, & \text{otherwise}
\end{cases}
$$

(b) Marginal pdf of $Y$:

Similarly, we integrate over $x$ from $0$ to $1 - y$:

$$
f_Y(y) = \int_0^{1 - y} 24xy dx = 24y \int_0^{1 - y} x dx
= 24y \cdot \left[ \frac{1}{2} x^2 \right]_0^{1 - y}
= 12y(1 - y)^2
$$

for $0 \leq y \leq 1$. Thus,

$$
f_Y(y) =
\begin{cases}
12y(1 - y)^2, & 0 \leq y \leq 1 \\
0, & \text{otherwise}
\end{cases}
$$
:::
::::

Figure \@ref(fig:toy-marginals) shows the marginal pdfs of $X$ and $Y$ from Examples \@ref(exm:chap-joint-continuous-toys) and \@ref(exm:chap-joint-toys-marginal). 

```{r toy-marginals, echo=FALSE, message=FALSE, fig.cap = "Marginal PDFs for Toy Example"}
x_vals <- seq(0, 1, length.out = 500)
f_x <- 12 * x_vals * (1 - x_vals)^2

y_vals <- seq(0, 1, length.out = 500)
f_y <- 12 * y_vals * (1 - y_vals)^2

par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))

plot(x_vals, f_x, type = "l", lwd = 2, col = "black",
     xlab = "x", ylab = expression(f[X](x)),
     main = "Marginal pdf of X")

plot(y_vals, f_y, type = "l", lwd = 2, col = "black",
     xlab = "y", ylab = expression(f[Y](y)),
     main = "Marginal pdf of Y")
```

Note that in this special case the two marginal pdfs are the same due to the symmetric structure of the joint pdf $f_{X,Y}$.  This won't in general be the case.

### More Than Two Random Variables

So far, we have considered only bivariate scenarios (two jointly-distributed random variables). However, the concepts of joint and marginal distributions generalize naturally to more than two random variables.

:::: {.whitebox data-latex=””}
::: {.example #chap-joint-three-vars}
Let $X$, $Y$, and $Z$ be continuous random variables with joint pdf
$$
f_{X,Y,Z}(x, y, z) =
\begin{cases}
8xyz, & \text{if } 0 \leq x \leq 1,\ 0 \leq y \leq 1,\ 0 \leq z \leq 1 \ \\
0, & \text{otherwise}
\end{cases}
$$

(a) Verify that $f_{X,Y,Z}(x, y, z)$ is a valid pdf.

(b) Compute $P(X \leq 0.5,\ Y \leq 0.5,\ Z \leq 0.5)$.

(c) Find the marginal pdf of $X$.

_Solution_:

(a) We integrate over the unit cube:

\begin{align*}
\int_0^1 \int_0^1 \int_0^1 &8xyz \, dz \, dy \, dx
= \int_0^1 \int_0^1 8xy \left[ \frac{1}{2} z^2 \right]_0^1 \, dy \, dx
= \int_0^1 \int_0^1 4xy \, dy \, dx \\
&= \int_0^1 4x \left[ \frac{1}{2} y^2 \right]_0^1 \, dx
= \int_0^1 2x \, dx = \left[ x^2 \right]_0^1 = 1
\end{align*}

So, $f_{X,Y,Z}(x, y, z)$ is a valid pdf.

(b) 

\begin{align*}
P(X \leq 0.5,\ &Y \leq 0.5,\ Z \leq 0.5) = \int_0^{0.5} \int_0^{0.5} \int_0^{0.5} 8xyz \, dz \, dy \, dx \\
&= \int_0^{0.5} \int_0^{0.5} 8xy \left[ \frac{1}{2} z^2 \right]_0^{0.5} \, dy \, dx 
= \int_0^{0.5} \int_0^{0.5} 8xy \cdot \left( \frac{1}{2} \right)^3 \, dy \, dx \\
&= \int_0^{0.5} \int_0^{0.5} xy \, dy \, dx 
= \int_0^{0.5} x \left[ \frac{1}{2} y^2 \right]_0^{0.5} \, dx \\
&= \int_0^{0.5} x \cdot \left( \frac{1}{2} \right)^3 \, dx 
= \int_0^{0.5} \frac{1}{8} x \, dx \\
&= \left[ \frac{1}{16} x^2 \right]_0^{0.5} = \frac{1}{16} \cdot \left( \frac{1}{2} \right)^2 
= \frac{1}{64}
\end{align*}

(c) We integrate out $Y$ and $Z$:

\begin{align*}
f_X(x) &= \int_0^1 \int_0^1 f_{X,Y,Z}(x, y, z) \, dy \, dz
= \int_0^1 \int_0^1 8xyz \, dy \, dz \\
&= \int_0^1 8xz \left[ \frac{1}{2} y^2 \right]_0^1 \, dz
= \int_0^1 4xz \, dz = 4x \left[ \frac{1}{2} z^2 \right]_0^1 = 2x
\end{align*}

So, the marginal pdf of $X$ is:

$$
f_X(x) =
\begin{cases}
2x, & 0 \leq x \leq 1 \\
0, & \text{otherwise}
\end{cases}
$$

:::
::::

## Independence

We introduced the concept of independence in Chapter \@ref(chap-probability), where we defined two events $A$ and $B$ to be independent if $P(A \cap B) = P(A)P(B)$. Heuristically, independence is always about a _lack_ of information.  Hence, knowing one event occurred should not affect the probability of the other. For example, in Example \@ref(exm:chap-probability-weather), we considered the events $A = {\text{it rains}}$ and $B = {\text{the forecaster predicts rain}}$. These events were not independent, since knowing the forecast influences the probability of rain. 

In this section, we revisit independence from the perspective of random variables. We will explore what it means for two (or more) random variables to be independent and how to determine independence based on their joint and marginal distributions.

:::: {.colbox data-latex=""}
::: {.definition #chap-joint-independence}
We say that the rvs $X$ and $Y$ are **independent** if

$$
f_{X,Y}(x, y) = f_X(x) \cdot f_Y(y) \quad \text{for all } x, y.
$$
This same definition applies regardless of if $f_{X,Y}$ is a pmf (for discrete rvs) or a pdf (for continuous rvs). 

If $X$ and $Y$ are not independent, then they are said to be **dependent**.  Note that being independent is a unique condition; there is only one way in which two random variables can be independent.  Conversely, there are unlimited ways in which two random variables can be dependent. We will discuss these ways (e.g. **correlation**) later in this chapter.
:::
::::

:::: {.whitebox data-latex=""}
::: {.example #chap-joint-coffee-shop-independence}
In Example \@ref(exm:chap-joint-coffee-shop), we described the joint behavior of two discrete random variables:

- $X$: the number of espresso shots ordered (0, 1, or 2), and  
- $Y$: whether the customer also buys a pastry ($0 =$ no, $1 =$ yes).

The joint pmf is given in Example \@ref(exm:chap-joint-coffee-shop), and we derived the marginal distributions in Example \@ref(exm:chap-joint-coffee-shop-marginal).

Are $X$ and $Y$ independent?

_Solution_:

To determine whether $X$ and $Y$ are independent, we must check whether
$$
f_{X,Y}(x, y) = f_X(x) \cdot f_Y(y) \quad \text{for all } x \text{ and } y.
$$

Let’s compute $f_X(x) \cdot f_Y(y)$ for a few combinations and compare it to $f_{X,Y}(x, y)$:

- $f_X(0)f_Y(0) = 0.15 \cdot 0.40 = 0.06 \ne f(0, 0) = 0.05$
- $f_X(1)f_Y(1) = 0.45 \cdot 0.60 = 0.27 \ne f(1, 1) = 0.25$
- $f_X(2)f_Y(0) = 0.40 \cdot 0.40 = 0.16 \ne f(2, 0) = 0.15$

Since $f_{X,Y}(x, y) \ne f_X(x) \cdot f_Y(y)$ for at least one $(x, y)$ pair (in fact, this equality fails to hold for several pairs), we conclude that $X$ and $Y$ are **dependent**.
:::
::::

:::: {.whitebox data-latex=""}
::: {.example #chap-joint-toys-independence}
In Example \@ref(exm:chap-joint-continuous-toys), we defined the joint pdf of $X$ and $Y$ as
$$
f_{X,Y}(x, y) =
\begin{cases}
24xy, & \text{if } 0 \le x \le 1,\ 0 \le y \le 1,\ \text{and } x + y \le 1 \\
0, & \text{otherwise}
\end{cases}
$$

Are $X$ and $Y$ independent?

_Solution_:

Recall that, in Example \@ref(exm:chap-joint-toys-marginal), we derived the marginal pdfs:
$$
f_X(x) = 
\begin{cases}
12x(1 - x)^2, & 0 \le x \le 1 \\
0, & \text{otherwise}
\end{cases}
\quad \text{and} \quad
f_Y(y) = 
\begin{cases}
12y(1 - y)^2, & 0 \le y \le 1 \\
0, & \text{otherwise}
\end{cases}
$$

To assess independence, we must check whether
$$
f_{X,Y}(x, y) = f_X(x) \cdot f_Y(y)
\quad \text{for all } (x, y) \text{ in the support}.
$$

Let us compare both sides for a general point $(x, y)$ such that $x + y \le 1$. The left-hand side is:
$$
f_{X,Y}(x, y) = 24xy.
$$

The right-hand side is:
$$
f_X(x) \cdot f_Y(y) = \left[ 12x(1 - x)^2 \right] \cdot \left[ 12y(1 - y)^2 \right] = 144xy(1 - x)^2(1 - y)^2.
$$

Clearly, $f_{X,Y}(x, y) \ne f_X(x) \cdot f_Y(y)$ unless both $(1 - x)^2 = 1$ and $(1 - y)^2 = 1$, which is only true at $x = 0$ and $y = 0$. Therefore, the factorization does not hold on the support in general. Conclusion: $X$ and $Y$ are *dependent*.
:::
::::

It is important to note that while independence is a strong assumption and often does *not* hold in observational settings (as we saw in the previous examples), there are important contexts in which assuming independence is reasonable. For instance, in a randomized experiment (recall definitions of observational studies and experiments in Section \@ref(chap-foundations-observational-experiment)), treatment assignment is carried out independently of any subject characteristics, so we may assume independence between the treatment indicator and baseline covariates. Similarly, in other real-world scenarios, independence can follow from the inherent structure of the variables being studied — for example, if $X$ is a randomly chosen subject’s birth month and $Y$ is their eye color, and if the population is well-mixed demographically, then it may be reasonable to assume that $X$ and $Y$ are independent.

The concept of independence extends naturally to collections of more than two random variables.

:::: {.colbox data-latex=""}
::: {.definition}
Let $X_1, X_2, \ldots, X_V$ be random variables (discrete or continuous). These variables are **mutually independent** if, for every subset $\{X_{v_1}, X_{v_2}, \ldots, X_{v_k}\} \subseteq \{X_1, X_2, \ldots, X_V\}$, the joint probability function (pmf or pdf) factorizes as

$$
f(x_{v_1}, x_{v_2}, \ldots, x_{v_k}) = f_{X_{v_1}}(x_{v_1}) f_{X_{v_2}}(x_{v_2}) \cdots f_{X_{v_k}}(x_{v_k})
$$

for all $x_{v_1}, x_{v_2}, \ldots, x_{v_k}$.
:::
::::

## [Conditional Probability Functions]{#def-joint-and-conditional-distributions-conditional-probability-function}

We introduced the idea of conditional probability in Chapter \@ref(chap-probability), where we explored how the probability of an event can change when we know that another event has occurred. For example, in Example \@ref(exm:chap-probability-weather), we saw that the probability of rain given a forecast predicting rain differs from the unconditional probability of rain. Similarly, when working with random variables, we often wish to understand the distribution of one variable **given that** we know the value of another. This leads us to the concept of a **conditional distribution**, which allows us to isolate and analyze the behavior of a random variable under specific conditions.

:::: {.colbox data-latex=""}
::: {.definition #conditional-distribution}


The **conditionally**  probability function (either pmf for discrete or pdf for continuous rvs) of $Y$ given $X$ is

$$
f_{Y|X}(y \mid x) = \frac{f_{X,Y}(x, y)}{f_X(x)}
$$

for all $(x,y)$ such that $f_X(x)) > 0$, where $f_X(x)$ is the marginal of $X$.


The conditional distribution describes the behavior of $Y$ when the value of $X$ is known.  Note that without loss of generality we can reverse the roles of $X$ and $Y$ in this definition.
:::
::::

:::: {.whitebox data-latex=""}
::: {.example #chap-joint-coffee-shop-conditional}
In the context of Example \@ref(exm:chap-joint-coffee-shop), recall that $X$ is the number of espresso shots ordered (0, 1, or 2) and $Y$ indicates whether the customer also buys a pastry ($Y = 0$ = no, $Y = 1$ = yes). Compute the probability that a customer orders 2 espresso shots, conditional on whether the customer did or did not buy a pastry. 

_Solution_:

First, recall that the marginal pmf of $Y$ is:

$$
\begin{aligned}
f_Y(0) &= 0.05 + 0.20 + 0.15 = 0.40 \\
f_Y(1) &= 0.10 + 0.25 + 0.25 = 0.60
\end{aligned}
$$

The conditional pmf of $X$ given $Y = 0$ satisfies:

$$
f_{X|Y}(x \mid y = 0) = \frac{f_{X,Y}(x, 0)}{f_Y(0)}
$$

Thus,

$$
f_{X|Y}(2 \mid y = 0) = \frac{f_{X,Y}(2, 0)}{f_Y(0)} = \frac{0.15}{0.40} = 0.375
$$

Similarly, the conditional pmf of $X$ given $Y = 1$ is:

$$
f_{X|Y}(x \mid y = 1) = \frac{f_{X,Y}(x, 1)}{f_Y(1)}
$$

Thus,

$$
f_{X|Y}(2 \mid y = 1) = \frac{0.25}{0.60} \approx 0.4167
$$

Recall that the marginal probability that a customer orders 2 espresso shots (regardless of pastry purchase) is:

$$
f_X(2) = f_{X,Y}(2, 0) + f_{X,Y}(2, 1) = 0.15 + 0.25 = 0.40
$$

Thus, knowing whether a customer buys a pastry changes the probability of ordering 2 shots:

- If a customer does not buy a pastry, the probability they order 2 shots is 0.375.
- If a customer does buy a pastry, the probability they order 2 shots is approximately 0.4167.
- Marginally (ignoring pastry purchase), the probability is 0.40.

This illustrates that the number of espresso shots ordered and whether a pastry is purchased are not independent.
:::
::::

:::: {.whitebox data-latex=""}
::: {.example #chap-joint-toys-conditional}
Recall the toy manufacturer context of Example \@ref(exm:chap-joint-continuous-toys), with $X$ representing blue pigment proportion and $Y$ representing red pigment proportion. Compute the conditional probability $P(X > 0.5 \mid Y < 0.5)$.

_Solution_:

First, observe that:

- The event $\{Y < 0.5\}$ corresponds to the region where $0\leq y < 0.5$.
- Furthermore, $x$ can range from $0$ to $1 - y$: $0 \leq x \leq 1-y$.

The conditional probability can be written as

$$
P(X > 0.5 \mid Y < 0.5) = \frac{P(X > 0.5 \cap Y < 0.5)}{P(Y < 0.5)}
$$

We already computed $P(X > 0.5\cap Y < 0.5)$ in Example \@ref(exm:chap-joint-continuous-toys):

$$
P(X > 0.5 \cap \ Y < 0.5) = 0.3125
$$

Now, compute $P(Y < 0.5)$. To do so, note that:

$$
P(Y < 0.5) = \int_0^0.5 \int_0^{1-y} 24xy \, dx \, dy
$$

Evaluate the inner integral first:

$$
\int_0^{1 - y} 24xy \, dx = 24y \left[ \frac{1}{2}x^2 \right]_0^{1 - y} = 12y(1 - y)^2
$$

Thus,

\begin{align*}
P(Y < 0.5) &= \int_0^{0.5} 12y(1 - y)^2 \, dy \\
&= 12 \int_0^{0.5} (y - 2y^2 + y^3) \, dy \\
&= 12 \left[ \frac{1}{2}y^2 - \frac{2}{3}y^3 + \frac{1}{4}y^4 \right]_0^{0.5}\approx 0.6875
\end{align*}

Finally, the conditional probability is

$$
P(X > 0.5 \mid Y < 0.5) \approx \frac{0.3125}{0.6875} = 0.4545
$$

Thus, given that the red pigment proportion is less than 50%, there is about a 45.45% chance that the blue pigment proportion exceeds 50%.
:::
::::

## Conditional Expected Values {#sec-conditional-expectation}

We have seen how the behavior of one rv can be characterized in terms of conditional probabilities involving another variable. In a similar spirit, it is often useful to compute the expected value of one random variable given that another variable takes a specific value. This leads to the concept of a **[conditional expected value]{#def-joint-and-conditional-distributions-conditional-expected-value}**, which captures the average behavior of a variable under the condition that certain information is known.

:::: {.colbox data-latex=""}
::: {.definition #chap-joint-conditional-expectation}
Let $X$ and $Y$ be two rvs.

The **conditional expected value of $Y$ given $X = x$**, denoted $\mu_{Y|X}(x) = E[Y \mid X = x]$, is defined as follows:

- **Discrete Case**: If $X$ and $Y$ are discrete random variables with conditional pmf $f_{Y|X}(y \mid x)$, then

$$
E[Y \mid X = x] = \sum_y y \, f_{Y|X}(y \mid x)
$$

- **Continuous Case**: If $X$ and $Y$ are continuous random variables with conditional pdf $f_{Y|X}(y \mid x)$, then

$$
E[Y \mid X = x] = \int_{-\infty}^{\infty} y \, f_{Y|X}(y \mid x) \, dy
$$

That is, $E[Y \mid X = x]$ represents the expected value of $Y$ when $X$ is known to equal $x$.
:::
::::

:::: {.whitebox data-latex=""}
::: {.example #chap-joint-coffee-shop-cond-ev}
Returning to the context of Example \@ref(exm:chap-joint-coffee-shop), let us compute the conditional expected value of $X$ (number of espresso shots) given $Y$ (whether or not a pastry was ordered).  Just like for conditional probability functions, we can exchange the roles of $X$ and $Y$ and define $E[X | Y = y]$ 

_Solution_:

(a) Compute $E[X | Y = 0]$.

First, we find the conditional pmf of $X$ given $Y = 0$:

\begin{align*}
f_{X|Y}(0|0) = P(X = 0 \mid Y = 0) &= \frac{f_{X,Y}(0,0)}{f_Y(0)} = \frac{0.05}{0.40} = 0.125 \\
f_{X|Y}(1|0) = P(X = 1 \mid Y = 0) &= \frac{f_{X,Y}(1,0)}{f_Y(0)} = \frac{0.20}{0.40} = 0.5 \\
f_{X|Y}(2|0) = P(X = 2 \mid Y = 0) &= \frac{f_{X,Y}(2,0)}{f_Y(0)} = \frac{0.15}{0.40} = 0.375
\end{align*}

Then, using the definition of conditional expected value:

$$
E[X \mid Y = 0] = (0)(0.125) + (1)(0.5) + (2)(0.375) = 0 + 0.5 + 0.75 = 1.25
$$

Thus, the expected number of espresso shots ordered by a customer who does not buy a pastry is 1.25.

(b) Compute $E[X \mid Y = 1]$.

Similarly, we find the conditional pmf of $X$ given $Y = 1$:

\begin{align*}
f_{X|Y}(0|1) = P(X = 0 \mid Y = 1) &= \frac{f_{X,Y}(0,1)}{f_Y(1)} = \frac{0.10}{0.60} \approx 0.1667  \\
f_{X|Y}(1|1) = P(X = 1 \mid Y = 1) &= \frac{f_{X,Y}(1,1)}{f_Y(1)} = \frac{0.25}{0.60} \approx 0.4167  \\
f_{X|Y}(2|1) = P(X = 2 \mid Y = 1) &= \frac{f_{X,Y}(2,1)}{f_Y(1)} = \frac{0.25}{0.60} \approx 0.4167 
\end{align*}

Then:

$$
E[X \mid Y = 1] \approx (0)(0.1667) + (1)(0.4167) + (2)(0.4167) = 1.25
$$

Thus, the expected number of espresso shots ordered by a customer who does buy a pastry is also 1.25. In this case, $E[X \mid Y = 0] = E[X \mid Y = 1] = 1.25$, meaning that the expected number of espresso shots ordered is the same regardless of whether a pastry is purchased. This is an implication of $X$ and $Y$ being independent, which we proved earlier.

It is *not generally true* that $E[X \mid Y = y]$ will be the same for all values of $y$. 
:::
:::: 

:::: {.whitebox data-latex=””}
::: {.example #chap-joint-toys-cond-ev}
Continuing with the toy manufacturer context from Example \@ref(exm:chap-joint-continuous-toys), let us compute the conditional expected value of $X$ (the blue pigment proportion) given that $Y = y$ (the red pigment proportion).

_Solution_:

Recall from Example \@ref(exm:chap-joint-toys-marginal) that the marginal pdf of $Y$ is:

$$
f_Y(y) =
\begin{cases}
12y(1 - y)^2, & 0 \leq y \leq 1 \\
0, & \text{otherwise}
\end{cases}
$$

The conditional pdf of $X$ given $Y = y$ is:

$$
f_{X|Y}(x \mid y) = \frac{f_{X,Y}(x, y)}{f_Y(y)}
\quad \text{for } 0 \leq x \leq 1 - y
$$

Thus:

$$
f_{X|Y}(x \mid y) = \frac{24xy}{12y(1 - y)^2} = \frac{2x}{(1 - y)^2}
\quad \text{for } 0 \leq x \leq 1 - y
$$

The conditional expected value $E[X \mid Y = y]$ is then:

$$
E[X \mid Y = y] = \int_0^{1 - y} x \, f_{X|Y}(x \mid y) \, dx
= \int_0^{1 - y} x \cdot \frac{2x}{(1 - y)^2} \, dx
= \frac{2}{(1 - y)^2} \int_0^{1 - y} x^2 \, dx
$$

Compute the integral:

$$
\int_0^{1 - y} x^2 \, dx = \left[ \frac{1}{3} x^3 \right]_0^{1 - y} = \frac{(1 - y)^3}{3}
$$

Thus:

$$
E[X \mid Y = y] = \frac{2}{(1 - y)^2} \cdot \frac{(1 - y)^3}{3}
= \frac{2}{3}(1 - y)
\quad \text{for } 0 \leq y \leq 1
$$

:::
::::

Figure \@ref(fig:toy-cond-ev) plots the conditional expected value $E[X \mid Y = y] = \frac{2}{3}(1 - y)$, $0 \leq y \leq 1$, as a function of $y.$ We observe that the expected proportion of blue pigment decreases linearly as the proportion of red pigment increases.

```{r toy-cond-ev, echo=FALSE, message=FALSE, fig.cap = "Conditional Expected Value from Toy Example"}
# Create sequence of y-values
y_vals <- seq(0, 1, length.out = 500)

# Compute corresponding conditional expected values
e_x_given_y <- (2/3) * (1 - y_vals)

# Plot
plot(
  y_vals, e_x_given_y, type = "l", lwd = 2, col = "black",
  xlab = "y (Proportion of Red Pigment)", 
  ylab = "E[X  |  Y = y]",#expression(E[X ~ "|" ~ Y == y]),
  main = "",
  ylim = c(0, max(e_x_given_y))
)
```

## Covariance and Correlation {#sec-cov-and-corr}

We have just spent some time detailing what it means for two variables to be independent.  Now, we will start to describe some ways in which variables can be dependent.  The most famous is a term you have likely encountered in everyday conversation, **correlation**, where it is often used to describe a relationship between two things. 

In rough terms, correlation measures both the strength and the direction of the association between two variables. By "association", we mean the tendency for one variable to change in a predictable way as the other variable changes. In this section, we will formalize these ideas by introducing two related concepts — **covariance** and **correlation** — that allow us to quantify the extent and nature of such co-relationships in a precise, mathematical way.

Before defining covariance and correlation, it is helpful to formalize how to compute the [expected value of a general function]{#def-joint-and-conditional-distributions-expected-value-of-a-general-function} of two jointly distributed random variables.

:::: {.colbox data-latex=""}
::: {.proposition #expected-value-hxy}
Let $X$ and $Y$ be jointly distributed rvs, and let $h(X, Y)$ be a function of $(X, Y)$. Then:

- Discrete Case: If $X$ and $Y$ are discrete random variables with joint pmf $f_{X,Y}(x, y)$, the expected value of $h(X, Y)$ is

$$
E\left[ h(X, Y) \right] = \sum_x \sum_y h(x, y) \, f_{X,Y}(x, y)
$$

- Continuous Case: If $X$ and $Y$ are continuous random variables with joint pdf $f_{X,Y}(x, y)$, the expected value of $h(X, Y)$ is

$$
E\left[ h(X, Y) \right] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} h(x, y) \, f_{X,Y}(x, y) \, dy \, dx
$$

In either case, the expected value of $h(X, Y)$ is obtained by taking a probability-weighted average of $h(x, y)$ across all possible pairs $(x, y)$.
:::
::::

### Covariance

:::: {.colbox data-latex=""}
::: {.definition #covariance-definition}
Let $X$ and $Y$ be jointly distributed random variables. The **covariance** between $X$ and $Y$, denoted $\text{Cov}(X, Y)$, measures the tendency of $X$ and $Y$ to be simultaneously abnormally large or small, on average. It is defined as follows:

$$
\text{Cov}[X, Y] = E\left[ (X - E[X])(Y - E[Y]) \right] 
$$
This can be computed using the previous result via the function $h(x,y) = (x - E[X])(y - E[Y])$
<!--
- **Discrete Case**: If $X$ and $Y$ are discrete random variables with joint pmf $p(x, y)$, then

$$
\text{Cov}(X, Y) = E\left[ (X - E[X])(Y - E[Y]) \right] = \sum_x \sum_y (x - E[X])(y - E[Y]) \, p(x, y)
$$

- **Continuous Case**: If $X$ and $Y$ are continuous random variables with joint pdf $f(x, y)$, then

$$
\text{Cov}(X, Y) = E\left[ (X - E[X])(Y - E[Y]) \right] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} (x - E[X])(y - E[Y]) \, f(x, y) \, dy \, dx
$$


The covariance $\text{Cov}[X, Y]$ is positive if larger values of $X$ tend to occur with larger values of $Y$, negative if larger values of $X$ tend to occur with smaller values of $Y$, and near zero if there is no consistent linear association between $X$ and $Y$.-->
:::
::::

Covariance provides a way to quantify how two variables move together. If $X$ tends to be above its mean at the same times that $Y$ is above its mean (and below its mean when $Y$ is also below its mean), the product $(X - E[X])(Y - E[Y])$ tends to be positive, resulting in a positive covariance. Conversely, if $X$ tends to be above its mean when $Y$ is below its mean (and vice versa), the product tends to be negative, leading to a negative covariance. If there is no systematic relationship between how $X$ and $Y$ deviate from their means, the positive and negative contributions tend to cancel out, and the covariance is near zero.

```{r cov-toy-scatters, echo=FALSE, message=FALSE, fig.cap = "Scatterplots for Four Toy Datasets"}
set.seed(456)  

# Sample size
n <- 30

# Case 1: Strong positive correlation
x1 <- runif(n, -2, 2)
y1 <- 2*x1 + rnorm(n, sd = 0.5)

# Case 2: Strong negative correlation
x2 <- runif(n, -2, 2)
y2 <- -2*x2 + rnorm(n, sd = 0.5)

# Case 3: Near-zero correlation
x3 <- runif(n, -2, 2)
y3 <- runif(n, -2, 2)

# Case 4: Non-linear, near-zero correlation
x4 <- runif(n, -2, 2)
y4 <- x4 ^ 2 + rnorm(n, sd = 0.5)

par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(x1, y1, pch = 19, col = "black",
     main = "Case 1: Strong Positive",
     xlab = "X", ylab = "Y", xlim = c(-3, 3), ylim = c(-6, 6))
abline(0, 0, lty = 2)
plot(x2, y2, pch = 19, col = "black",
     main = "Case 2: Strong Negative",
     xlab = "X", ylab = "Y", xlim = c(-3, 3), ylim = c(-6, 6))
abline(0, 0, lty = 2)
plot(x3, y3, pch = 19, col = "black",
     main = "Case 3: Near Zero",
     xlab = "X", ylab = "Y", xlim = c(-3, 3), ylim = c(-6, 6))
abline(0, 0, lty = 2)
plot(x4, y4, pch = 19, col = "black",
     main = "Case 4: Near Zero (Non-Linear)",
     xlab = "X", ylab = "Y", xlim = c(-3, 3), ylim = c(-6, 6))
abline(0, 0, lty = 2)
```

Figure \@ref(fig:cov-toy-scatters) shows scatterplots of four toy datasets. Case 1 illustrates strong positive correlation, case 2 illustrates strong negative correlation, and case 3 illustrates near-zero correlation. Notice that in each of cases 1-3, the $X$ and $Y$ variables are _linearly_ related to one another. In case 4, on the other hand, the $X$ and $Y$ variables are _non-linearly_ related to one another; in particular, $Y$ was generated using a quadratic function of $X$. While $X$ and $Y$ are clearly _associated_ with one another (as $X$ changes values, $Y$ also changes in a systematic way), they are not _linearly_ associated. Covariance is a measure of _linear_ association, and hence it is not an appropriate measure of association for variables that are not linearly associated. Interestingly, while there is a clear systematic relationship between $X$ and $Y$ in case 3, their covariance is approximately zero. 

It is possible to reduce the number of calculations required to compute a covariance with the following result.

:::: {.colbox data-latex=""}
::: {.proposition #covariance-shortcut}
$$
\text{Cov}[X, Y] = E[XY] - E[X] \cdot E[Y] = E[XY] - \mu_X \cdot \mu_Y
$$
:::
::::

:::: {.whitebox data-latex=””}
::: {.example #chap-joint-coffee-shop-covariance}
Continuing with the coffee shop context from Example \@ref(exm:chap-joint-coffee-shop), let us compute the covariance between $X$ (number of espresso shots) and $Y$ (pastry purchase indicator). We will make use of the shortcut formula from Proposition \@ref(prp:covariance-shortcut).

_Solution_:

First, we compute $E[XY]$ using the joint pmf and the function $h(x,y) = xy$:

$$
\begin{align*}
E[XY] &= \sum_x \sum_y x y \cdot f_{X,Y}(x, y) \\
&= (0)(0.05) + (0)(0.10) + (0)(0.20) + (1)(0.25) + (0)(0.15) + (2)(0.25) \\
&= 0 + 0 + 0 + 0.25 + 0 + 0.50 = 0.75
\end{align*}
$$
We now compute $\mu_X$ and $\mu_Y$ using the marginal pmfs:

$$
\begin{aligned}
E[X] = \mu_X &= (0)(0.15) + (1)(0.45) + (2)(0.40) = 0 + 0.45 + 0.80 = 1.25 \\
E[Y] = \mu_Y &= (0)(0.40) + (1)(0.60) = 0 + 0.60 = 0.60
\end{aligned}
$$
We then have

\begin{align*}
\text{Cov}[X, Y] &= E[XY] - \mu_X \cdot \mu_Y \\
&= 0.75 - 1.25 \cdot 0.60 = 0
\end{align*}

While $X$ and $Y$ are not independent, they happen to have zero covariance. This is consistent with our finding in Example \@ref(exm:chap-joint-coffee-shop-cond-ev) that the conditional expected value $E[X \mid Y = y]$ equals the marginal expected value $E[X] = \mu_X$. It is possible for dependent rvs to have zero covariance.

It is however necessary for independent rvs to have zero covariance.
:::
::::

:::: {.whitebox data-latex=””}
::: {.example #chap-joint-toys-covariance}
Continuing with the toy manufacturing context from Example \@ref(exm:chap-joint-continuous-toys), let us compute the covariance between $X$ (proportion of blue pigment) and $Y$ (proportion of red pigment). 

_Solution_:

We first compute $E[XY]$ by integrating the joint pdf over the triangular support $0 \le x \le 1$, $0 \le y \le 1 - x$:

$$
\begin{aligned}
E[XY] &= \int_0^1 \int_0^{1 - x} x y \cdot f_{X,Y}(x, y) \, dy \, dx \\
&= \int_0^1 \int_0^{1 - x} x y \cdot 24xy\, dy\, dx \\
&= \int_0^1 \int_0^{1 - x} 24x^2 y^2 \, dy \, dx \\
&= \int_0^1 24x^2 \left[ \frac{1}{3} y^3 \right]_0^{1 - x} \, dx \\
&= \int_0^1 8x^2 (1 - x)^3 \, dx \\
&= \frac{2}{15}
\end{aligned}
$$
We now compute $\mu_X$ and $\mu_Y$ by integrating the marginal pdfs:

$$
\mu_X = \int_0^1 x \cdot f_X(x) \, dx = \int_0^1 12x^2(1 - x)^2 \, dx = \frac{2}{5}
$$
By symmetry, we also have $\mu_Y = \frac{2}{5}$. Thus,

$$
\text{Cov}[X, Y] = \frac{2}{15} - \left( \frac{2}{5} \cdot \frac{2}{5} \right) = \frac{2}{15} - \frac{4}{25} = -\frac{2}{75}
$$
The covariance between $X$ and $Y$ is $-2/75$, indicating a negative linear association between the proportion of blue pigment and red pigment in the toy formula.
:::
::::

Note that a significant defect in the definition (and hence usage) of covariance is that the value $-\frac{2}{75}$ is essentially meaningless.  The sign (negative in this case) is meaningful, demonstrating a negative association.  But the magnitude is in terms of the units of $X$ times the units of $Y$, which is almost always uninterpretable.  Essentially, it is impossible to say if $X$ and $Y$ are *strongly* or *weakly* associated based on the number $-\frac{2}{75}$.  For that, we need covariance's more famous cousin: **correlation**.

### Correlation

Covariance provides a useful way to assess the direction and strength of a _linear_ relationship between two random variables. However, because covariance is measured in units of $X$ times units of $Y$, its magnitude can be difficult to interpret or compare across different settings. For this reason, a related concept — **correlation** — is often preferred in practice. Correlation rescales covariance to a unitless quantity between $-1$ and $1$, making it easier to interpret as it conveys information about the strength and direction of the relationship.  For this reason correlation is more common in statistical reports. A correlation of $+1$ or $-1$ indicates a perfect linear relationship, while a correlation near $0$ suggests little to no linear association. Like covariance, correlation reflects only _linear_ relationships and may not capture nonlinear dependencies.

:::: {.colbox data-latex=””}
::: {.definition #chap-joint-correlation}
Let $X$ and $Y$ be jointly distributed random variables. The correlation between $X$ and $Y$, denoted $\text{Cor}[X, Y]$, is defined as:

$$
\rho_{X,Y} = \text{Cor}[X, Y] = \frac{\text{Cov}[X, Y]}{\sigma_X \cdot \sigma_Y}
$$

where $\sigma_X = \sqrt{\text{Var}[X]}$ and $\sigma_Y = \sqrt{\text{Var}[Y]}$ are the standard deviations of $X$ and $Y$, respectively.

The correlation $\rho_{X,Y}$ always lies between $-1$ and $1$ and is a unitless measure of the strength and direction of the linear relationship between $X$ and $Y$.

- $\rho_{X,Y} = 1$ implies a perfect positive linear relationship.
- $\rho_{X,Y} = -1$ implies a perfect negative linear relationship.
- $\rho_{X,Y} = 0$ implies no linear relationship (though a nonlinear relationship may still exist).
:::
::::

Recall the toy example datasets shown in Figure \@ref(fig:cov-toy-scatters). The correlation between the two variables would be close to 1 for case 1, close to $-1$ for case 2, and close to $0$ for cases 3 and 4. 

:::: {.whitebox data-latex=””}
::: {.example #chap-joint-coffee-shop-correlation}
Returning to the coffee shop example from Example \@ref(exm:chap-joint-coffee-shop), let us compute the correlation between $X$ (number of espresso shots) and $Y$ (pastry purchase indicator). 

_Solution_: We previously computed:

- $\text{Cov}[X, Y] = 0$
- $\mu_X = 1.25$, and $\mu_Y = 0.60$

Since the covariance equals zero, it is clear that the correlation will also equal zero. Nevertheless, we illustrate the calculation. Using the marginal pmfs, we can easily compute the standard deviations $\sigma_X = \sqrt{0.4875} \approx 0.698$ and $\sigma_Y = \sqrt{0.2400} \approx 0.490$. Thus,

\begin{align*}
\rho_{X,Y} &= \frac{\text{Cov}[X, Y]}{\sigma_X \cdot \sigma_Y} \\
&\approx \frac{0}{0.698 \cdot 0.490} = 0
\end{align*}

As we have previously noted, in this example, while $X$ and $Y$ are not independent, they do not have a systematic linear association.
:::
::::

Again, covariance and correlation are measures of _linear_ association between variables. In particular, it is possible for two dependent rvs to have covariance and correlation values of zero. 

:::: {.colbox data-latex=""}
::: {.proposition #covariance-shortcut}
If two rvs $X$ and $Y$ are independent, then their correlation (and covariance) is zero. However, the converse is not necessarily true: $\rho_{X,Y} = 0$ does **not** imply that $X$ and $Y$ are independent.
:::
::::

:::: {.whitebox data-latex=””}
::: {.example #chap-joint-toys-correlation}
Returning to the toy manufacturing context from Example \@ref(exm:chap-joint-continuous-toys), let us compute the correlation between $X$ (proportion of blue pigment) and $Y$ (proportion of red pigment).  Intuitively, there would be a relationship between the pigment levels as there is a resource constraint; having more blue necessarily means having less red.  

_Solution_: We previously computed:

- $\text{Cov}[X, Y] = -\frac{2}{75}$
- $\mu_X = \mu_Y = \frac{2}{5}$

We can obtain the standard deviations $\sigma_X$ and $\sigma_Y$ by integrating the marginal pdfs, leading us to $\sigma_X = \sigma_Y = 1/5$. Thus,

$$
\begin{align*}
\rho_{X,Y} &= \frac{\text{Cov}[X, Y]}{\sigma_X \cdot \sigma_Y} \\
&= -\frac{\frac{2}{75}}{\frac{1}{5} \cdot \frac{1}{5}} = -\frac{(2) (25)}{75} = -\frac{2}{3}
\end{align*}
$$
Consistent with what we saw from computing the covariance, the correlation indicates that there is a moderate _negative_ linear association between the proportions of blue and red pigments. 
:::
::::

The following proposition states a couple of useful facts about correlation.

:::: {.colbox data-latex=""}
::: {.proposition #corr-properties}
Let $X$ and $Y$ be rvs, and let $a$, $b$, $c$, and $d$ be real-valued constants.

1. The correlation between linearly transformed variables satisfies  
$$
\text{Cor}[aX + b,\ cY + d] = \text{Cor}[X, Y]
$$

2. Correlation is always bounded between $-1$ and $1$:
$$
-1 \leq \rho_{X,Y} \leq 1
$$
Moreover,
$$
\rho_{X,Y} = \pm 1 \quad \text{if and only if} \quad Y = aX + b \text{ for some constants } a \neq 0,\, b.
$$
:::
::::

The first result tells us that correlation is unaffected by linear transformations of the variables. In particular, changing the units of measurement—such as converting from Fahrenheit to Celsius—does not change the correlation between two variables.

For example, suppose $X$ and $Y$ represent daily high temperatures (in degrees Fahrenheit) recorded in two different cities. If $\rho_{X,Y}$ denotes the correlation between them, then the correlation between the corresponding Celsius measurements,
$$
X^* = \frac{5}{9}(X - 32) \quad \text{and} \quad Y^* = \frac{5}{9}(Y - 32),
$$
is exactly the same: $\text{Cor}[X^*, Y^*] = \rho_{X,Y}$.

Finally, remember that **correlation is not the same as causation**. Even a strong correlation between two variables does not imply that changes in one _cause_ changes in the other. As discussed in Section \@ref(chap-foundations-observational-experiment), we can only make causal claims when our data arise from a well-designed experiment. In contrast, data from observational studies can reveal associations but leave open the possibility that the observed relationship is due to a **lurking** or **confounding** variable—some unobserved factor that influences both $X$ and $Y$.
